State of the Practice for Medical Imaging Software

### 2021-11-17 StateOfPractice/Papers/MedImageSoft_SOP/MedImageSoft_SOP_AcademicRadiology.tex
Latest edits to the MI State of the Practice paper

@@

@@

\title{Assessing the State of the Practice for Medical Imaging Software}

@@

\begin{abstract}
%% Text of abstract

@@

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword

%% PACS codes here, in the form: \PACS code \sep code

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

@@

%% main text
\section{}
\label{}

>>>>

\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{mindmap}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{arydshln}

@@

\pgfplotsset{compat=1.13}

@@

\title{State of the Practice for Medical Imaging Software}

@@

\begin{abstract}

We present a general method to assess the state of the practice for Scientific
Computing (SC) software and apply this method to Medical Imaging (MI) software.
We selected 29 MI software projects from 48 candidates, assessed 10 software
qualities (installability, correctness/ verifiability, reliability, robustness,
usability, maintainability, reusability, understandability,
visibility/transparency and reproducibility) by answering 103 questions for each
software project, and interviewed 8 of the 29 development teams. Based on the
quantitative data for the first 9 qualities, we ranked the MI software with the
Analytic Hierarchy Process (AHP). The top three software products were
\textit{3D Slicer}, \textit{ImageJ}, and \textit{OHIF Viewer}. By interviewing
the developers, we identified three major pain points: i) lack of resources; ii)
difficulty balancing between compatibility, maintainability, performance, and
security; and, iii) lack of access to real-world datasets for testing. For
future MI software projects, we propose adopting test-driven development, using
continuous integration and continuous delivery (CI/CD), using git and GitHub,
maintaining good documentation, supporting third-party plugins or extensions,
considering web application solutions, and improving collaboration between
different MI software projects.

@@

\begin{keyword}
	medical imaging, scientific computing, software engineering, software
	quality, Analytic Hierarchy Process, developer interviews
\end{keyword}

@@

\section{Introduction} \label{ch_intro}

We aim to study the state of software development practice for Medical Imaging
(MI) software.  MI tools use images of the interior of the body (from sources
such as Magnetic Resonance Imaging (MRI), Computed Tomography (CT), Positron
Emission Tomography (PET) and Ultrasound) to provide information for diagnostic,
analytic, and medical applications \citep{FDA2021, enwiki:1034887445,
Zhang2008}.  Given the importance of MI software and the high number of
competing software projects, we wish to understand the merits and drawbacks of
the current development processes, tools and methodologies.  We aim to assess
through a software engineering lens the quality of the existing software with
the hope of highlighting standout examples, understanding current pain points
and providing guidelines and recommendations for future development.

\subsection{Purpose} \label{sec_motivation}

Not only do we wish to gain insight into the state of the practice for MI
software, we also wish to understand the development of research software in
general. We wish to understand the impact of the often cited gap, or chasm,
between software engineering and scientific programming \citep{Storer2017}.
Although scientists spend a substantial proportion of their working hours on
software development \citep{Hannay2009, Prabhu2011}, many developers learn
software engineering skills by themselves or from their peers, instead of from
proper training \citep{Hannay2009}. \citet{Hannay2009} observe that many
scientists showed ignorance and indifference to standard software engineering
concepts. For instance, according to a survey by \citet{Prabhu2011}, more than
half of their 114 subjects did not use any proper debugger for their software.

To gain insights, we devised five research questions, which can be applied
to MI, and to other domains, of research software \citep{SmithEtAl2021}:

\begin{description}

\item[RQ1.] \hypertarget{rq1}What artifacts are present in current software
projects? What role does documentation play in the projects? What are the
developers' attitude toward it?
\item[RQ2.] \hypertarget{rq2}What tools are used in the development of current
software packages?
\item[RQ3.] \hypertarget{rq3}What principles, processes, and methodologies are
used in the development of current software packages?
\item[RQ4.] \hypertarget{rq4}What are the pain points for developers working on
research software projects? What aspects of the existing processes,
methodologies, and tools do they consider as potentially needing improvement?
What changes to processes, methodologies, and tools can improve software
development and software quality?
\item[RQ5.] \hypertarget{rq5}What is the current status of the following
software qualities for the projects? What actions have the developers taken to
address them?
\begin{itemize}
	\item Installability
	\item Correctness \& Verifiability
	\item Reliability
	\item Robustness
	\item Usability
	\item Maintainability
	\item Reusability
	\item Understandability
	\item Visibility/Transparency
	\item Reproducibility
\end{itemize}
\item[RQ6.] \hypertarget{rq6}How does software designated as high quality by
this methodology compare with top-rated software by the community?
\end{description}

\subsection{Overview of the Methodology}

We designed a general methodology to assess the state of the practice for SC
software \citep{SmithEtAl2021}. Details can be found in Section~\ref{}.  This
methodology builds off prior work to assess the state of the practice for such
domains as Geographic Information Systems \citep{smith2018state}, Mesh
Generators \citep{smith2016state}, Seismology software
\citep{Smith2018Seismology}, and Statistical software for psychology
\citep{smith2018statistical}.  In keeping with the previous methodology, we have
maintained the constraint that the work load for measuring a given domain should
be feasible for a team as small as one person, and for a short time,
ideally around a person month of effort. A person month is considered to be $20$
working days ($4$ weeks in a month, with $5$ days of work per week) at $8$
person hours per day, or $20 \cdot 8 = 160$ person hours.

With our methodology, we first choose an SC domain (in the current case MI) and
identify a list of about 30 software packages. We approximately measure the
qualities of each package by filling in a grading template (forward ref?).
Compared with our previous methodology, the new methodology also includes
repository based metrics, such as the number of files, number of lines of code,
percentage of issues that are closed, etc. (forward ref?).  With the
quantitative data in the grading template, we rank the software with the
Analytic Hierarchy Process (AHP) (forward ref?). After this, as another addition
to our previous methodology, we interview some of the development teams to
further understand the status of their development process (forward ref?).
Finally, we summarize the results and propose recommendations for future SC
software development (forward ref?).

\subsection{Scope} \label{sec_scope}

We need to restrict the scope of MI software that we consider so that the
measurement exercise is feasible within a person month of effort. MI considers a
broad range of different problems.  According to \citet{Bankman2000}, MI
software deals with six different basic problems, while \citet{Angenent2006}
pointed out that four fundamental problems are solved by MI software. While both
mentioned Segmentation, Registration, and Visualization of medical images,
Bankman also included Enhancement, Quantification, and three functions for MI
archiving and telemedicine systems (Compression, Storage, and Communication)
\citep{Bankman2000}. On the other hand, \citet{Angenent2006} included
Simulation. According to Wikipedia contributors \citep{enwiki:1034877594}, MI
software has primary functions in Segmentation, Registration, Visualization
(including the basic display, reformatted views, and 3D volume rendering),
Statistical Analysis, and Image-based Physiological Modelling. As
\citet{Kim2011} describes, the general steps of medical image analysis after
obtaining digital data include Enhancement, Segmentation, Feature Extraction,
Classification, and Interpretation. Besides the above major functions, some MI
software provides supportive functions. For example, with Tool Kit libraries VTK
\citep{SchroederEtAl2006} and ITK \citep{McCormick2014}, developers build
software with Visualization and Analysis functions; Picture Archiving and
Communication System (PACS) helps users to economically store and conveniently
access images \citep{Choplin1992}.

Based on our literature survey, we divided MI software into five sub-groups and
several sub-sub-groups by their major functions, as shown in Figure
\ref{fig_mi_functions}.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[mindmap, grow cyclic, every node/.style=concept, concept
color=orange!40,
level 1/.append style={level distance=5cm,sibling angle=72},
level 2/.append style={level distance=3cm,sibling angle=90},
level 3/.style={level distance=2.3cm,sibling angle=90}]
\node{MI software}
child[concept color=teal!40] { node {Enhancement}}
child[concept color=teal!40] { node {Analysis}
    child[concept color=blue!30] { node {Registration}}
    child[concept color=blue!30,] { node {Segmentation}}
    child[concept color=blue!30] { node {Statistical Analysis}
    child[concept color=blue!20] { node {Feature Extraction}}
    child[concept color=blue!20] { node {Classification}}
    child[concept color=blue!20] { node {Interpretation}}
    }}
child[concept color=teal!40] { node {Simulation/\\Modeling}}
child[concept color=teal!40] { node {Supporting}
    child[concept color=blue!30] { node {Tool Kit}}
    child[concept color=blue!30] { node {PACS}}
}
child[concept color=teal!40] { node {Visualization}
    child[concept color=blue!30] { node {2D Display}}
    child[concept color=blue!30] { node {3D Rendering}}
    child[concept color=blue!30] { node {Reformatted Views}}
};
\end{tikzpicture}
\caption{Major functions of MI software}
\label{fig_mi_functions}
\end{figure}

To keep the data collection and analysis feasible, we limited the scope of the
software to the software packages providing the Visualization tools and
functions in this project.

[Roadmap - instead sprinkle forward references throughout the introduction?]

\section{Background} \label{ch_background}

When designing a method for evaluating the state of the practice of
domain-specific software, we included a step to select domain and software.
Knowledge of different software categories is essential for the selection. To
compare and rank the software qualities with the grading template in Appendix
\ref{ap_grading_template}, we need the definitions of qualities and the AHP.

In this section, we introduce the relevant software categories (Section
\ref{sec_software_categories}). We also cover the software quality definitions
(Section \ref{sec_software_quality}) and an overview of the AHP (Section
\ref{sec_AHP}).

\subsection{Software Categories} \label{sec_software_categories}

We target specific software categories to narrow down the scope when selecting
software packages for measuring. One way to categorize them is grouping by
functions, such as analyzing the scope for the MI software (Section
\ref{sec_scope}), which requires specific domain knowledge. In this section, we
discuss three common software categories that we can apply to all SC domains:
Open source software (OSS), freeware, and commercial software.

\subsubsection{Open Source Software} \label{sec_open_source_software}

For OSS, the source code is openly accessible. Users have the right to study,
change and distribute it under a license granted by the copyright holder. For
many OSS projects, the development process relies on the collaboration of
different contributors worldwide \citep{Corbly2014}. Accessible source code
usually exposes more ``secrets'' of a software project, such as the underlying
logic of software functions, how developers achieve their works, and the flaws
and potential risks in the final product. Thus, it brings much more convenience
to researchers analyzing the qualities of the project.

\subsubsection{Freeware} \label{sec_freeware}

Freeware is software that can be used free of charge. Unlike with OSS, the
authors of freeware do not allow users to access or modify the source code of
the software \citep{LINFO2006}. The term \textit{freeware} should not be confused
with \textit{free software}, which is similar to OSS. To the end-users, the
differences between freeware and OSS often do not bother them. The fact that
these products are free of charge is likely to make them popular with many
users. However, software developers, end-users who wish to modify the source
code, and researchers looking for insight into software development process will
find the inaccessible source code a problem.

\subsubsection{Commercial Software}

``Commercial software is software developed by a business as part of its
business'' \citep{GNU2019}. Typically speaking, the users are required to pay to
access all of the features of commercial software, excluding access to the
source code. However, some commercial software is also free of charge
\citep{GNU2019}. Based on our experience, most commercial software products are
not OSS.

For some specific software, the backgrounds of commercial software developers
often differ from the ones of non-commercial OSS. In such a case, the former is
usually the product of software engineers, and the latter is likely to have
developers who work in the domain and are also end-users of the products. One
example of such software is SC software, since the developers need to utilize
their domain-specific during the development process \citep{Wilson2014}.

\subsection{Software Quality Definitions} \label{sec_software_quality}

This section lists the definitions of 10 software qualities, which are from
Smith et al. \citep{SmithEtAl2020}. We aim to measure each of them for selected
SC software packages. The order of the first nine qualities follows our grading
template in Appendix \ref{ap_grading_template}. We do not measure
\textit{reproducibility} with the grading template, but discuss it with the
developers by interviews.

\begin{itemize}
	\item \textbf{Installability} The effort required for the installation,
	uninstallation, or reinstallation of a software or product in a specified
	environment \citep{ISO/IEC25010} \citep{lenhard2013measuring}.

	\item \textbf{Correctness \& Verifiability} A program is correct if it
	behaves according to its stated specifications. Verifiability is the extent
	to which a set of tests can be written and executed, to demonstrate that the
	delivered system meets the specification \citep{GhezziEtAl2003}.

	\item \textbf{Reliability} The probability of failure-free operation of a
	computer program in a specified environment for a specified time, i.e., the
	average time interval between two failures also known as the mean time to
	failure (MTTF) \citep{musa1987software} \citep{GhezziEtAl2003}.

	\item \textbf{Robustness} Software possesses the characteristic of
	robustness if it behaves ``reasonably'' in two situations: i) when it
	encounters circumstances not anticipated in the requirements specification,
	and ii) when the assumptions in its requirements specification are violated
	\citep{ghezzi1991fundamentals} \citep{boehm2007software}.

	\item \textbf{Usability} ``The extent to which a product can be used by
	specified users to achieve specified goals with effectiveness, efficiency,
	and satisfaction in a specified context of use" \citep{ISO/TR16982:2002}
	\citep{ISO9241-11:2018}.

	\item \textbf{Maintainability} The effort with which a software system or
	component can be modified to i) correct faults; ii) improve performance or
	other attributes; iii) satisfy new requirements
	\citep{IEEEStdGlossarySET1990} \citep{boehm2007software}.

	\item \textbf{Reusability} ``The extent to which a software component can be
	used with or without adaptation in a problem solution other than the one for
	which it was originally developed" \citep{kalagiakos2003non}.

	\item \textbf{Understandability} ``The capability of the software product to
	enable the user to understand whether the software is suitable, and how it
	can be used for particular tasks and conditions of use" \citep{iso2001iec}.

	\item \textbf{Visibility/Transparency} The extent to which all of the steps
	of a software development process and the current status of it are conveyed
	clearly \citep{ghezzi1991fundamentals}.

	\item \textbf{Reproducibility} ``A result is said to be reproducible if
	another researcher can take the original code and input data, execute it,
	and re-obtain the same result" \citep{BenureauAndRougier2017}.
\end{itemize}

\subsection{Analytic Hierarchy Process} \label{sec_AHP}

To generate ranking scores for a set of software packages, we use AHP, which
utilizes pairwise comparisons between all of the packages. Thomas L. Saaty
developed this tool, and people widely used it to make and analyze multiple
criteria decisions \citep{VaidyaEtAl2006}. AHP organizes multiple criteria
factors in a hierarchical structure and uses pairwise comparisons between
alternatives to calculate relative ratios \citep{Saaty1990}.

For a project with $ m $ criteria, we can use an $m\times m$ matrix $A$ to
record the relative importance between factors. When comparing criterion $i$ and
criterion $j$, the value of $A_{ij}$ is decided as follows, and the value of
$A_{ji}$ is $1/A_{ij}$ \citep{Saaty1990},

\begin{itemize}
	\item $A_{ij} = 1$ if criterion $i$ and criterion $j$ are equally important;
	\item $A_{ij} = 9$ if criterion $i$ is extremely more important than
	criterion $j$;
	\item $A_{ij}$ equals to an integer value between 1 and 9 according the the
	relative importance of criterion $i$ and criterion $j$.
\end{itemize}

The above process assumes that criterion $i$ is not less important than
criterion $j$, otherwise, we need to reverse $i$ and $j$ and determine $A_{ji}$
first, then $A_{ij} = 1/A_{ji}$.

The priority vector $w$ can be calculated by solving the following equation
\citep{Saaty1990}, 
\begin{equation} 
    Aw = \lambda_{max}w,
\end{equation}
where $\lambda_{max}$ is the maximal eigenvalue of $A$.

In this project, $w$ is approximated with the classic \textit{mean of normalized
values} approach \citep{AlessioEtAl2006},

\begin{equation}
w_i = \frac{1}{m}\sum_{j=1}^{m}\frac{A_{ij}}{\sum_{k=1}^{m}A_{kj}}
\end{equation}

Suppose there are $n$ alternatives, for criterion $i = 1, 2, ... , m$, we can
create an $n\times n$ matrix $B_i$ to record the relative preferences between
these choices. The way of generating $B_i$ is similar to the one for $A$.
However, unlike comparing the importance between criteria, we pairwise decide
how much we favor one alternative over the other. We use the same method to
calculate the local priority vector for each $B_i$.

In this project, the first nine software qualities mentioned in Section
\ref{sec_software_quality} are the criteria ($m = 9$), while 29 software
packages ($n = 29$) are compared for each of the $m$ criteria. The software are
evaluated with the grading template in Appendix \ref{ap_grading_template} and a
subjective score from one to ten is given for each quality for each package. For
each quality, for a pair of packages $i$ and $j$, such that $score_i >=
score_j$, the difference between two scores is $\mathit{diff_{ij}} = score_i -
score_j$. The relationship between $A_{ij} = 1$ and $\mathit{diff_{ij}}$ is as
follows:

\begin{itemize}
\item $A_{ij} = 1$ and $\mathit{diff_{ij}} = 0$ when criterion $i$ and criterion
$j$ are equally important;
\item $A_{ij}$ increases when $\mathit{diff_{ij}}$ increases;
\item $A_{ij} = 9$ and $\mathit{diff_{ij}} = 9$ when criterion $i$ is extremely
more important than criterion $j$.
\end{itemize}

Thus, we approximate the pairwise comparison result of $i$ versus $j$ by the
following equation:

\begin{equation}
A_{ij} = min(score_i - score_j + 1, 9)
\end{equation}

\section{Methodology} \label{ch_methods}

We designed a general process for evaluating the state of the practice of
domain-specific SC software, that we instantiate for a specific scientific
domain.

Our method involves four steps:
\begin{enumerate}
\item choosing a software domain (Section \ref{sec_domain_selection});
\item collecting and filtering software packages (Section
\ref{sec_software_selection});
\item grading the selected software (Section \ref{sec_grading_software});
\item interviewing development teams for further information (Section
\ref{sec_interview_methods}).
\end{enumerate}

\noindent Section \ref{sec_applying_method} presents an example of how we
applied the method on the MI domain.

\subsection{Domain Selection} \label{sec_domain_selection}

Our methods are generic for any SC software, but they need to be applied to a
specific domain. When choosing a candidate domain, we prefer one with a large
number of active OSS. The reason is that we aim to finalize a list of 30
software packages \citep{SmithEtAl2021} after the screening step. For example, we
remove the ones without recent updates or specific functions. Besides, we prefer
OSS projects because our grading method requires access to the code. In
addition, we prefer a domain with an active community developing and using the
software. As a result, it is easier to invite enough developers for interviews.

We prefer 30 software packages providing similar functions or falling into
different sub-groups depending on our research purpose. So the domain needs to
have enough candidates in one sub-group or enough sub-groups to cross-compare.

We also prefer domains in which our team has expertise. We invite domain experts
to join and support our projects. They help us in many aspects, such as vetting
the software list and interview questions.

\subsection{Software Product Selection} \label{sec_software_selection}

The process of selecting software packages contains two steps: i) identify
software candidates in the chosen domain, ii) filter the list according to our
needs \citep{SmithEtAl2021}.

\subsubsection{Identify Software Candidates} \label{sec_identify_software_candidates}

We start with finding candidate software in publications of the domain. Then, we
search various websites, such as \hyperlink{https://github.com/}{GitHub},
\hyperlink{https://swmath.org/}{swMATH} and the Google search results for
software recommendation articles. We should also include the ones suggested by
the domain experts \citep{SmithEtAl2021}.

\subsubsection{Filter the Software List} \label{sec_filter_software_list}

The goal is to build a software list with a length of about 30
\citep{SmithEtAl2021}.

The only ``mandatory'' requirement is that the software must be OSS, as defined
in Section \ref{sec_open_source_software}. We need this because evaluating some
software qualities requires the source code.

The other filters are optional, and we consider them according to the number of
software candidates and the objectives of the research project. We apply them in
the following priority order:

\begin{enumerate}
\item The functions and purpose of the software. An SC domain often contains
software with various functions and purposes. For example, some MI software
packages are Tool Kit for developers to use, and some others offer Visualization
function to end-users. We have two options:
\begin{itemize}
\item selecting a set of software with the same major function. In this case, we
can use the identical process to assess all packages, e.g., same input to
measure \textit{Robustness}. Also, when we give impression scores to qualities
such as \textit{Installability} and \textit{Usability}, the results are more
comparable. Thus, it is more feasible to collect the results and rank the
software.
\item selecting software from a set of different sub-groups. For example, we can
choose 10 MI software from each of the three sub-groups: Visualization, Tool
Kit, and PACS. The downside: we may need different processes to measure each
sub-group; it may be less accurate to mix all three sub-groups and rank them
together. The benefit: we can cross-compare the development processes between
the sub-groups.
\end{itemize}

\item The version control tool. The empirical measurement tools listed in
Section \ref{sec_empirical_measurements} only work on projects using
\hyperlink{https://git-scm.com/}{Git}, so we prefer software with Git. Some
manual steps in empirical measurement depend on a few metrics of GitHub, which
makes projects held on GitHub more favored \citep{SmithEtAl2021}.

\item The age of the software. Some of the OSS projects may experience a lack of
recent maintenance. So we eliminate packages without recent updates, unless they
are still popular and highly recommended by the domain users
\citep{SmithEtAl2021}. We consider a software project as ``alive" if it has any
update within the last 18 months; otherwise, we mark it as ``dead".
\end{enumerate}

The order of filters 2 and 3 is flexible. We adjust it according to the number
of software packages affected by the filters, and the number of ones remaining
on the list.

\subsubsection{Vet the Software List} \label{sec_vet_software_list}

Before showing our filtered list to the domain experts, we ask them to list
their top 10 software in the domain. Then, we cross-compare the two lists and
discuss the commonality and variability.  In addition, we ask them to vet our
filtered list. They provide views on whether the list is reasonable. We also use
their opinions for a filtering process. For example, if a software package is
not OSS and has had no updates for a long time, but the domain experts identify
it as a valuable product, we still consider it in our final list.

\subsection{Grading Software} \label{sec_grading_software}

We grade the selected software using a template (Section
\ref{sec_grading_template}) and a specific empirical method (Section
\ref{sec_empirical_measurements}). Some technical details for the measurements
are in Section \ref{sec_technical_details}.

\subsubsection{Grading Template} \label{sec_grading_template}

The full grading template can be found in Appendix \ref{ap_grading_template}.
The template contains 103 questions that we use for grading software products.
Figure \ref{fg_grading_template_example} shows an example of this grading
template.

\begin{figure}[ht]
\includegraphics[scale=0.42]{figures/template.png}
\caption{Grading template example}
\label{fg_grading_template_example}
\end{figure}

We use the first section of the template to collect general information, such as
the name, purpose, platform, programming language, publications about the
software, the first release and the most recent change date, website, source
code repository of the product, etc. Information in this section helps us
understand the projects better and may be helpful for further analysis, but it
does not directly affect the grading scores.

We designed the following nine sections in the template for the nine software
qualities mentioned in Section~\ref{sec_software_quality}. For each quality, we
ask several questions and the typical answers are among the collection of
``yes'', ``no'', ``n/a'', ``unclear'', a number, a string, a date, a set of
strings, etc. Each quality needs an overall score between 1 and 10 based on all
the previous questions. For some qualities, we perform surface measurements,
which allow us to measure all packages with reasonable efforts. The surface
measurements reveal some traits of a underlying quality, but may not fully
represent it.

\begin{itemize}

\item \textbf{Installability} We check the existence and quality of installation
instructions. The user experience is also an important factor, such as the ease
to follow the instructions, number of steps, automation tools, and the
prerequisite steps for the installation. If any problem interrupts the process
of installation or uninstallation, we give a lower score to this quality. We
also record the Operating System (OS) for the installation test and whether we
can verify the installation.

\item \textbf{Correctness \& Verifiability} For \textit{correctness}, we check
the projects to identify the techniques to ensure this quality, such as literate
programming, automated testing, symbolic execution, model checking, unit tests,
etc. We also examine whether the projects use continuous integration and
continuous delivery (CI/CD). For \textit{verifiability}, we go through the
documents of the projects to check the requirements specifications, theory
manuals, and getting started tutorials. If a getting started tutorial exists and
provides expected results, we follow it and check if the outputs match.

\item \textbf{Surface Reliability} We check whether the software breaks during
the installation and the operations following a getting started tutorials,
whether there are descriptive error messages, and if we can recover the process
after an error. We use the software to load damaged images during the assessment
to \textit{surface robustness}, and lower its score for \textit{surface
reliability} if a software product ``break" in this process.

\item \textbf{Surface Robustness} We check how the software handles
unexpected/unanticipated input. For example, we prepare broken image files for
MI software packages with the function to load image files. We use a text file
(.txt) with a modified extension name (.dcm) as an unexpected/unanticipated
input. We load a few correct input files to ensure the function is working
correctly before testing with the unexpected/unanticipated ones.

\item \textbf{Surface Usability} We examine the documentation of the projects,
and we consider software with a getting started tutorial and a user manual
easier to use. Meanwhile, we check if users have any channels to get support. We
also record our impressions and user experiences when testing the software.
Easy-to-use graphical user interfaces give us a better experience, which leads
to better scores.

\item \textbf{Maintainability} We search the projects' documents and identify
the process of contributing and reviewing code. We believe that the artifacts of
a project - including source code, documents, building scripts, etc. - can
significantly affect its  \textit{maintainability}. Thus we check each project
for its artifacts, such as API documentation, bug tracker, release notes, test
cases, build files, version control, etc. We also check the tools supporting
issue tracking and version control, the percentages of closed issues, and the
proportion of comment lines in code.

\item \textbf{Reusability} We count the total number of code files for each
project. Projects with a large number of components provide more choices to
reuse. Furthermore, well-modularized code, which tends to have smaller parts in
separated files, is typically easier to reuse. Thus, we consider the projects
with more code files and less code lines per file to be more reusable. We use a
command-line tool \textit{scc} to count the number of  text-based files and LOC
for all projects. We also decide that the projects with API documentation can
deliver better \textit{Reusability}.

\item \textbf{Surface Understandability} We randomly examine 10 code files. We
check the code’s style within each file, such as whether the identifiers,
parameters, indentation, and formatting are consistent, whether the constants
(other than 0 and 1) are hardcoded, and whether the developers modularized the
code. We also check the descriptive information for the code, such as documents
mentioning the coding standard, the comments in the code, and the descriptions
or links for algorithms in the code.

\item \textbf{Visibility/Transparency} To measure this quality, we check the
existing documents to find out whether the software development process and
current status of a project are visible and transparent. We examine the
development process, current status, development environment, and release notes
for each project. If any information is missing or poorly conveyed, the
\textit{visibility/transparency} will be lower.
\end{itemize}

For some qualities, the empirical measurements also affect the score. We use
tools to extract information from the source code repositories. For projects
held on GitHub, we manually collect additional metrics, such as the stars of the
GitHub repository, and the numbers of open and closed pull requests. Section
\ref{sec_empirical_measurements} presents more details about the empirical
measurements.

\subsubsection{Empirical Measurements} \label{sec_empirical_measurements}

We use two command-line tools for the empirical measurements. One is
\textit{GitStats} that generates statistics for git repositories and displays
outputs in the form of web pages \citep{Gieniusz2019}; the other one is Sloc Cloc
and Code (as known as \textit{scc}) \citep{Boyter2021}, aiming to count the lines
of code, comments, etc.

Both tools measure the number of text-based files in a git repository and lines
of text in these files. Based on our experience, most text-based files in a
repository contain programming source code, and developers use them to compile
and build software products. A minority of these files are instructions and
other documents. So we roughly regard the lines of text in text-based files as
lines of programming code. The two tools usually generate similar but not
identical results. From our understanding, this minor difference is due to the
different techniques to detect if a file is text-based or binary.

Additionally, we manually collect information for projects held on GitHub, such
as the numbers of stars, forks, people watching this repository, open pull
requests, closed pull requests, and the number of months a repository has been
on GitHub. A git repository can have a creation date much earlier than the first
day on GitHub. For example, the developers created the git repository of
\textit{3D Slicer} in 2002, but did not upload a copy of it to GitHub until
2020. We get the creation date of the GitHub copy by using API
\textit{https://api.github.com/repos/{:owner}/{:repository}} (e.g.,
\hyperlink{https://api.github.com/repos/slicer/slicer}{https://api.github.com/repos/slicer/slicer}).
In the response, the value of ``created\_at" is what we want. The number of
months a repository has been on GitHub helps us understand the average change of
metrics over time, e.g., the average new stars per month.

These empirical measurements help us from two aspects. Firstly, they help us
with getting a project overview faster and more accurately. For example, the
number of commits over the last 12 months shows how active this project has
been, and the number of stars and forks may reveal its popularity. Secondly, the
results may affect our decisions regarding the grading scores for some software
qualities. For example, if the percentage of comment lines is low, we
double-check the \textit{understandability} of the code; if the ratio of open
versus closed pull requests is high, we pay more attention to the
\textit{maintainability}.

\subsubsection{Technical Details} \label{sec_technical_details}

To test the software on a ``clean'' system, we create a new virtual machine (VM)
for each software and only install the necessary dependencies before measuring.
We make all 30 VMs on the same computer, one at a time, and destroy them after
measuring.

We spend about two hours grading each package, unless we find technical issues
and need more time to resolve them. In most of the situation, we finish all the
measurements for one software on the same day.

\subsection{Interview Methods} \label{sec_interview_methods}

This section introduces our interview questions (Section
\ref{sec_interview_questions}), method of selecting interviewees (Section
\ref{sec_interviewee_selection}), and interview process (Section
\ref{sec_interview_process}).

\subsubsection{Interview Questions} \label{sec_interview_questions}

We designed a list of 20 questions to guide our interviews, which can be found
in Section \ref{ch_interview} and Appendix \ref{ap_interview}.

Some questions are about the background of the software, the development teams,
the interviewees, and how they organize the projects. We also ask about their
understandings of the users. Some questions focus on the current and past
difficulties, and the solutions the team has found or will try. We also discuss
the importance and current situations of documentation. A few questions are
about specific software qualities, such as \textit{maintainability},
\textit{understandability}, \textit{usability}, and \textit{reproducibility}.

The interviews are semi-structured based on the question list; we ask follow-up
questions when necessary. Based on our experience, the interviewees usually
bring up some exciting ideas that we did not expect, and it is worth expanding
on these topics.

\subsubsection{Interviewee Selection} \label{sec_interviewee_selection}

For a software list with a length of roughly 30, we aim to interview about ten
development teams. Interviewing multiple individuals from each team gives us
more comprehensive information, but a single engineer well-knowing the project
is also sufficient.

Ideally, we select projects after the grading measurements and prefer the ones
with higher overall scores. However, if we do not find enough participants, we
reach all teams on the list. As mentioned in Section
\ref{sec_apply_to_mi_interviews}, when we applied this process to the MI domain,
we eventually contacted all teams.

We try to find the contacts of the teams on the projects' websites, such as the
official web pages, repositories, publications, and bio pages of the teams'
institutions. Then, we send at most two emails to one contact asking for its
participation before receiving any replies. We operate the invitation according
to our ethics approval, such as the one in Appendix \ref{ap_ethics}. For
example, we ask for participants' consent before interviewing them, recording
the conversation, or including it in our report.

\subsubsection{Interview Process} \label{sec_interview_process}

Before contacting any interviewee candidate, we need to receive ethics clearance
from the McMaster University Research Ethics Board. Since the members of the
development teams are usually around the world, we organize these interviews as
virtual meetings online with \hyperlink{https://zoom.us/}{Zoom}. After receiving
consent from the interviewees, we also record and transcribe our discussions.

\subsection{Applying the Method to MI} \label{sec_applying_method}

This section shows an overview of applying our method to the MI domain.

\subsubsection{Domain Selection}

Based on the principles in Section \ref{sec_domain_selection}, we selected the
MI domain and the sub-group of software with the Visualization function shown in
Figure \ref{fig_mi_functions}. We also included Dr. Michael Noseworthy, a
professor of Electrical and Computer Engineering at McMaster University,
Co-Director of the McMaster School of Biomedical Engineering, and Director of
Medical Imaging Physics and Engineering at St. Joseph’s Healthcare, and some of
his students as the MI domain experts in our team.

\subsubsection{Software Product Selection} \label{sec_mi_software_selection}

By using the method in Section \ref{sec_identify_software_candidates}, we
identified 48 MI software projects as the candidates from publications
\citep{Bjorn2017} \citep{Bruhschwein2019} \citep{Haak2015}, online articles related
to the domain \citep{Emms2019} \citep{Hasan2020} \citep{Mu2019}, forum discussions
related to the domain \citep{Samala2014}, etc. Appendix
\ref{ap_list_before_filtering} shows all 48 software packages.

Guided by the method in Section \ref{sec_filter_software_list}, we filtered the
list with a process as follows:

\begin{enumerate}

\item Among them, there were eight that we could not find their source code,
such as \textit{MicroDicom}, \textit{Aliza}, and \textit{jivex}. These packages
are likely to be freeware defined in Section \ref{sec_freeware} and not OSS. So
following guidelines in Section \ref{sec_filter_software_list} we removed them
from the list.

\item Next, we focused on the MI software providing Visualization functions, as
described in Section \ref{sec_scope}. Seven of the software on the list were
Tool Kits or libraries for other software to use as dependencies, but not for
end-users to view medical images, such as \textit{VTK}, \textit{ITK}, and
\textit{dcm4che}; another three were PACS. We also eliminated these from the
list.

\item Finally, we removed \textit{Open Dicom Viewer} from the list because it
had not received any updates for a long time (since 2011). After that, only
\textit{MatrixUser} and \textit{AMIDE} were considered as ``dead". However, both
of them had much more recent updates (after 2017) than \textit{Open Dicom
Viewer}.

\end{enumerate}

We still preferred projects using git and GitHub and being updated recently, but
did not apply this filter since packages were already below 30. Even without
this filter, 27 out of the 29 software packages on the filtered list used git,
and 24 chose GitHub.

Following the process in Section \ref{sec_vet_software_list}, our domain experts
provided a list of top software that contains 12 software products (Table
\ref{tab_top_software_experts}). We compared two lists and found six common
ones.

\begin{table}[ht]
\centering
\begin{tabular}{ll}
\hline
Software & On both lists \\ \hline
3D Slicer & X \\
Horos & X \\
ImageJ & X \\
Fiji & X \\
AFNI &  \\
FSL &  \\
Freesurfer &  \\
Mricron & X \\
Mango & X \\
Tarquin &  \\
Diffusion Toolkit &  \\
MRItrix &  \\ \hline
\end{tabular}
\caption{\label{tab_top_software_experts}Top software by the MI domain experts}
\end{table}

We included \textit{Mango} in the initial list, but removed it because it was
not OSS. However, we kept \textit{Papaya}, a the web version of \textit{Mango}.
Instead of \textit{MRIcron}, we chose \textit{MRIcroGL}, because
\textit{MRIcron} development had moved to \textit{MRIcroGL} \citep{Rorden2021b}.

Six software packages on the domain experts' list were not on our filtered list.
We believed their primary function was Analysis mentioned in Section
\ref{sec_scope}. Thus, we did not include them in our final list.

After vetting our filtered list, the domain experts believed it was reasonable
and did not identify any problem. Thus, as shown in Appendix
\ref{ap_list_before_filtering}, eventually, we had 29 software products on the
final list.

\subsubsection{Grading Software}
\label{sec_applying_method_grading}
Then we followed the steps in Section \ref{sec_grading_software} to measure and grade the software. 27 out of the 29 packages are compatible with two or three different OS such as Windows, macOS, and Linux, and 5 of them are browser-based, making them platform-independent. However, in the interest of time, we only performed the measurements for each project by installing it on one of the platforms, most likely Windows.

\subsubsection{Interviews}
\label{sec_apply_to_mi_interviews}
We received ethics clearance from the McMaster University Research Ethics Board
(Appendix \ref{ap_ethics}). Going through the interview process in Section
\ref{sec_interview_methods}, we contacted all of the 29 teams. Members from
eight teams responded and agreed to participate. As a result, we interviewed
nine developers and architects from the eight teams.

\section{Measurement Results} \label{ch_results}
